The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-11-18 script/en/chapters/primes.tex
Script[de+en]: hanging indent in footnotes; corrected links to old appendix parts
Script[de+en]: Added current research about twin primes (chap 3.8.3); made some small layout changes

\newpage
\hypertarget{Kapitel_2}{}
\chapter{Prime Numbers}
\label{Label_Kapitel_2}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007, July 2009)

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Catalan's conjecture]{Catalan's conjecture\footnotemark}
    \footnotetext{%
Eugene Charles Catalan, Belgian mathematician, May 5, 1814$-$Feb 14, 1894.\\
After him the so-called {\em Catalan numbers}
$A(n) = (1 / (n+1) ) * (2n)! / (n!)^2 $ \\
$=  1, 2, 5, 14, 42, 132, 429, 1430, 
4862, 16796, 58786, 208012, 742900, 2674440, 9694845, ... $
are named.
    }
\nopagebreak
Catalan \index{Catalan Eugene}  \index{Number!Catalan}
conjectured that $ C_4 \;$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
\begin{sloppypar}
(see \url{http://www.utm.edu/research/primes/mersenne.shtml} under Conjectures and
Unsolved Problems).
\end{sloppypar}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{Open questions}
Twin primes are prime numbers whose difference is 2. 
Examples include 5 and 7 or 101 and 103. 
Triplet primes, however, only occur once: 3, 5, 7.  
For all other sets of three consecutive uneven numbers, one of them is 
always divisible by 3 and thus not a prime.

\begin{itemize}

\item The number of twin primes is an open question: infinite or limited
    number? The largest twin primes known today are 
    $1,693,965 \cdot 2^{66,443} \pm 1.$

\item Does a formula exist for calculating the number of twin primes per
    interval?

\item \hypertarget{link-Primzahlfunktion-base-a-Offen}{}
    The \hyperlink{link-Primzahlfunktion-base-a-Proof}{above} proof of the
    function $f: \: N \rightarrow Z$ with $n \mapsto a^{3^n}$  
    only guarantees the existence of such a number $a$.  How can we
    determine this number $a$ and will it have a value, making the 
    function also of some practical interest?

>>>>

\newpage
\hypertarget{Kapitel_2}{}
\chapter{Prime Numbers}
\label{Label_Kapitel_2}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007, November 2009)

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Catalan's conjecture]{Catalan's conjecture\footnotemark}
    \footnotetext{%
Eugene Charles Catalan, Belgian mathematician, May 5, 1814$-$Feb 14, 1894.\\
After him the so-called {\em Catalan numbers}
$A(n) = (1 / (n+1) ) * (2n)! / (n!)^2 $ \\
$=  1, 2, 5, 14, 42, 132, 429, 1430, 
4862, 16796, 58786, 208012, 742900, 2674440, 9694845, ... $
are named.
    }
\nopagebreak
Catalan \index{Catalan Eugene}  \index{Number!Catalan}
conjectured that $ C_4 \;$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
%% \begin{sloppypar}
\noindent (see \url{http://www.utm.edu/research/primes/mersenne.shtml}
           under ``Conjectures and Un\-sol\-ved Problems'').
%% \end{sloppypar}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{Open questions about twin primes}

Twin primes are prime numbers whose difference is 2. 
Examples include 5 and 7, or 101 and 103,
or $1,693,965 \cdot 2^{66,443} \pm 1.$.

\noindent The biggest known twin pair nowadays is \[65,516,468,355\cdot 2^{333,333} \pm 1\]
with $100,355$ decimal digits.\footnote{\url{http://primes.utm.edu}, October 2009}

\noindent Remark: Triplet primes, however, only occur once: 3, 5, 7.  
For all other sets of three consecutive uneven numbers, one of them is 
always divisible by 3 and thus not a prime.

\noindent Open questions are:
\begin{itemize}
\item The number of twin primes: infinitely many or a limited number?

\item Does a formula exist for calculating the number of twin primes per interval?
\end{itemize}

A big step towards the solution of this problem was made by Dan Goldston,
J\'{a}nos Pintz and Cem Yildirim in 2003\footnote{%
    D. A. Goldston: ``Gaps Between Primes''\\
    \url{http://www.math.sjsu.edu/~goldston/OberwolfachAbstract.pdf}\\ 
    See also: 
    \begin{itemize}
       \item D. A. Goldstone: "'Are There Infinitely Many Twin Primes?"\\
            \url{http://www.math.sjsu.edu/~goldston/twinprimes.pdf}
       \item K. Soundararajan: "'Small Gaps Between Prime Numbers:
            The Work Of Goldston-Pintz-Yildirim"\\
	    \url{http://www.ams.org/bull/2007-44-01/S0273-0979-06-01142-6/S0273-0979-06-01142-6.pdf}
    \end{itemize}
                                               }. 
The three mathematicians were investigating the distribution of prime numbers.
They could proof, that \[ \liminf_{n\to\infty}{\frac{p_{n+1}-p_n}{\log{p_n}}}=0,\]
where $p_{n}$ denotes the $n$-th prime number. \\
This means that the smallest limit point ($\liminf$) of the sequence
$\frac{p_{n+1}-p_n}{\log{p_n}}$ equals zero.\\
A point is called limit point of a sequence, if there lie in any arbitrarily small
neighbourhood of that point infinitely many elements of the sequence.\\
$\log{p_n}$ is about the average distance between the prime $p_n$ and the next prime
$p_{n+1}$.\\
Hence, the term above implies, that there are infinitely many consecutive
primes with a gap between them which is arbitrarily small compared to the expected
average gap.

Moreover, it was proofed, that \[p_{n+1}-p_n < (\log{p_n})^{8/9}\] holds true
for infinitely many primes\footnote{ c't magazine 2003, no. 8, page 54}.

Those results could be the basis for the proof, that infinitely many twin primes exist.

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{Further open questions}

\begin{itemize}
\item \hypertarget{link-Primzahlfunktion-base-a-Offen}{}
    The \hyperlink{link-Primzahlfunktion-base-a-Proof}{above} proof of the
    function $f: \: N \rightarrow Z$ with $n \mapsto a^{3^n}$  
    only guarantees the existence of such a number $a$.  How can we
    determine this number $a$ and will it have a value, making the 
    function also of some practical interest?
